---
title: "phase4+australia"
format: html
editor: visual
---

```{r}
# load libraries 
library(Seurat)
library(SeuratData)
library(SeuratWrappers)
library(ggplot2)
library(patchwork)
library(harmony)
library(qs)
library(tidyverse)
library(dplyr)
library(rstatix)
library(pheatmap)
library(Matrix)
library(reshape2)
library(biomaRt)
```

```{r}
# load filtered phase 4  
micro_AD_CO <- readRDS("micro_AD_CO.rds")
micro_AD_CO # already a seurat object 
dim(micro_AD_CO) # gene x cell 
DefaultAssay(micro_AD_CO) <- "RNA"
dim(micro_AD_CO) # gene x cell 
table(micro_AD_CO$orig.ident) 
length(table(micro_AD_CO$orig.ident)) # check length (71 samples/donors)
table(micro_AD_CO$Status) # check how many cells are AD and CO
unique(micro_AD_CO$Status) # check if there are only AD and CO
prop.table(table(micro_AD_CO$Status)) * 100 # percentage
table(micro_AD_CO$gender) # check how many cells are male and female 
prop.table(table(micro_AD_CO$gender)) * 100 # percentage 
table(micro_AD_CO$AOD) 
table(micro_AD_CO$Status2) # breakdown of AD (ADAD/sAD)
prop.table(table(micro_AD_CO$Status2)) * 100 # percentage 
tapply(micro_AD_CO$orig.ident, micro_AD_CO$Status2, function(x) length(unique(x))) # number of donors per sub status

sample_status <- tapply( # check number of donors per status (AD or CO)
  micro_AD_CO$Status,
  micro_AD_CO$orig.ident,
  unique
)
table(sample_status)

sample_info <- data.frame( # check number of disease status + sex per status 
  Sample = micro_AD_CO$orig.ident,
  Status = micro_AD_CO$Status,
  Sex = micro_AD_CO$gender
)
sample_info_unique <- unique(sample_info) # keep one row per donor
prop.table(table(sample_status)) * 100 # percentage 
table(sample_info_unique$Status, sample_info_unique$Sex) 
prop.table(table(sample_info_unique$Status, sample_info_unique$Sex), margin = 1) * 100 

# checking apoe donors 
unique(micro_AD_CO$orig.ident[micro_AD_CO$flag_APOE4 == "APOE4+"])
table(micro_AD_CO$orig.ident[micro_AD_CO$flag_APOE4 == "APOE4+"])
length(table(micro_AD_CO$orig.ident[micro_AD_CO$flag_APOE4 == "APOE4+"]))

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~#
# load preprocessed australia 
microglia.aus <- qread("microglia_reprocessed.qs") # AUS microglia dataset (originally 19626 cells)
microglia.aus # already a seurat object 
dim(microglia.aus) # gene x cell 
DefaultAssay(microglia.aus) <- "RNA"
dim(microglia.aus) # gene x cell 
table(microglia.aus$orig.ident) 
length(table(microglia.aus$orig.ident)) # check length (71 samples/donors)
table(microglia.aus$Status) # check how many cells are AD and CO
unique(microglia.aus$Status) # check if there are only AD and CO
prop.table(table(microglia.aus$Status)) * 100 # percentage
table(microglia.aus$Sex) # check how many cells are male and female 
prop.table(table(microglia.aus$Sex)) * 100 # percentage 
table(microglia.aus$Age_of_death) 
table(microglia.aus$Case_sub_status) # breakdown of AD (ADAD/sAD)
prop.table(table(microglia.aus$Case_sub_status)) * 100 # percentage 
tapply(microglia.aus$orig.ident, microglia.aus$Case_sub_status, function(x) length(unique(x))) # number of donors per sub status

sample_status <- tapply( # check number of donors per status (AD or CO)
  microglia.aus$Status,
  microglia.aus$orig.ident,
  unique
)
table(sample_status)

sample_info <- data.frame( # check number of disease status + sex per status 
  Sample = microglia.aus$orig.ident,
  Status = microglia.aus$Status,
  Sex = microglia.aus$Sex
)
sample_info_unique <- unique(sample_info) # keep one row per donor
prop.table(table(sample_status)) * 100 # percentage 
table(sample_info_unique$Status, sample_info_unique$Sex) 
prop.table(table(sample_info_unique$Status, sample_info_unique$Sex), margin = 1) * 100 # percentage

# checking apoe donors 
unique(microglia.aus$orig.ident[microglia.aus$APOE4_y_n == "Yes"])
table(microglia.aus$orig.ident[microglia.aus$APOE4_y_n == "Yes"])
length(table(microglia.aus$otablerig.ident[microglia.aus$APOE4_y_n == "Yes"]))

# cohort characterization table with mutation details (PSN1, APP) for ADAD samples
adad <- microglia.aus@meta.data[microglia.aus$Case_sub_status == "ADAD", ]

psen1_donors <- unique(adad$ID[adad$PSEN1_mutation_y_n == "Yes"])
psen1_donors
length(psen1_donors)

app_donors <- unique(adad$ID[adad$APP_mutation__y_n == "Yes"])
app_donors
length(app_donors)

data.frame( # table
  Mutation = c("PSEN1", "APP"),
  Donor_Count = c(length(psen1_donors), length(app_donors))
)
```

```{r}
DefaultAssay(microglia.aus) <- "RNA"
DefaultAssay(micro_AD_CO) <- "RNA"
# keep only to RNA assay and only the raw counts layer of the dataset 
microglia.aus <- DietSeurat(
  microglia.aus,
  assays = "RNA",
  counts = TRUE,
  data = FALSE,
  scale.data = FALSE
)
micro_AD_CO <- DietSeurat(
  micro_AD_CO,
  assays = "RNA",
  counts = TRUE,
  data = FALSE,
  scale.data = FALSE
)
```

Analysis of basis data. (cells, umis) The 'Idents' = 'seurat_clusters'
```{r}
splitted.data <- SplitObject(micro_AD_CO, split.by = "orig.ident") # splitting data by donors 
# remove these two samples that have very few cells
splitted.data$`46_16258_F` <- NULL
splitted.data$`5_1442_F` <- NULL

# number of features and cells in each sample - phase 4
map_dfr(splitted.data, function(x) {
  data.frame(
    sample = unique(x$orig.ident),
    ncells = ncol(x),
    ngenes = nrow(x),
    group = unique(x$Status)
  )
}) -> info

dims <- t(sapply(splitted.data, function(x) dim(x)))
sort(dims[,2])  

aus.split <- SplitObject(microglia.aus, split.by = "orig.ident") # splitting data by donors 
# number of features and cells in each sample - australia 
map_dfr(aus.split, function(x) {
  data.frame(
    sample = unique(x$orig.ident),
    ncells = ncol(x),
    ngenes = nrow(x),
    group = unique(x$Status)
  )
}) -> info

dims <- t(sapply(aus.split, function(x) dim(x)))
sort(dims[,2])  

aus.split$`MGI3522_CKAE-SU804-1` <- NULL # low cell count

VlnPlot(microglia.aus, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3) # checking amount of umis per cell 
VlnPlot(micro_AD_CO, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

micro_AD_CO$percent.mt <- pmin(micro_AD_CO$percent.mt, 10) # setting max to 10%
VlnPlot(micro_AD_CO, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

```{r}
# match genes - keeps only genes present in both 
common_genes <- intersect(
  rownames(microglia.aus),
  rownames(micro_AD_CO)
)
microglia.aus <- subset(microglia.aus, features = common_genes) # changed from 32547 genes to 31194 genes 
micro_AD_CO <- subset(micro_AD_CO, features = common_genes) # changed from 32569 genes to 31194 genes 
dim(microglia.aus)
dim(micro_AD_CO)
length(intersect(rownames(microglia.aus), rownames(micro_AD_CO))) # checking overlap - should correspond w/ genes 
```

checking class - class() checking structure - str()
```{r}
# merge objects - combines cells from both datasets 
combined <- merge(
  microglia.aus,
  y = micro_AD_CO,
  add.cell.ids = c("AUS", "ADCO"))
dim(combined) 
length(table(combined$orig.ident))
```

```{r}
# creating a metadata column that labels which dataset each cell came from
combined$batch <- ifelse(
  grepl("^AUS_", colnames(combined)),
  "AUS",
  "ADCO"
)
table(combined$batch) # sanity check 
```

```{r}
# split the single object into a list of objects by sample/identity
combined <- SplitObject(combined, split.by = "orig.ident")

# filter out objects whose RNA counts matrix has 0 genes
combined <- combined[sapply(combined, function(x)
    dim(GetAssayData(x, assay = "RNA", layer = "counts"))[1] > 0)]
```

# Renormalize data
```{r}
# run SCTransform on the list 
combined <- lapply(combined, function(x) {
    print(unique(x$orig.ident))
    SCTransform(
      x,
      vars.to.regress = c("nFeature_RNA", "subsets_Mito_percent", "subsets_Ribo_percent"),
      verbose = TRUE,
      return.only.var.genes = FALSE,
      variable.features.n = Inf
    )
})
class(combined) # check class 
```

# Get the list of protein coding genes 
```{r}
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
pc <- getBM(attributes = c("hgnc_symbol", "gene_biotype"),
            mart = ensembl) %>% 
  filter(gene_biotype == "protein_coding", hgnc_symbol != "") 
pc <- unique(pc$hgnc_symbol)
```

# Exclude non-coding genes
```{r}
all_var_features <- c()
# select only protein-coding variable genes and add them to all variable features vector
for (x in combined) {
    sample_var_genes <- x$SCT@var.features[x$SCT@var.features %in% pc]
    print(length(sample_var_genes))
    print(length(sample_var_genes))
    all_var_features <- union(all_var_features, sample_var_genes)
}
print(length(all_var_features))

# exclude genes that are not expressed in all datasets
# the limiting step to get a higher number of genes is this filtering
# in this case, 2300 genes are expressed across all samples
for (x in combined) {
    all_var_features <- all_var_features[all_var_features %in% rownames(x$SCT)]
}
print(length(all_var_features)) 
head(all_var_features)
```

```{r}
list_genes <- map(combined, function(x) { x$SCT@var.features[x$SCT@var.features %in% pc] })

# initialize matrix
samples <- names(list_genes)
n <- length(samples)
shared_matrix <- matrix(0, nrow = n, ncol = n, dimnames = list(samples, samples))

# Calculate pairwise proportion of shared proteins (Jaccard index)
for (i in seq_along(samples)) {
  for (j in seq_along(samples)) {
    a <- list_genes[[i]]
    b <- list_genes[[j]]
    shared_matrix[i, j] <- length(intersect(a, b)) / length(union(a, b))
  }
}

# plot
pheatmap(shared_matrix,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         main = "Proportion of shared proteins")
```

# Drop samples 13, 21, 27, 54 and 'MGI3522_CKAE-SU639-1' b/c they are outliers
```{r}
combined <- combined[-c(grep("^13_|^21_|^27_|^54_", names(combined)))]
combined <- combined[!names(combined) %in% "MGI3522_CKAE-SU639-1"]

all_var_features <- c() 
# select only protein-coding variable genes and add them to all variable features vector
for (x in combined) {
    sample_var_genes <- x$SCT@var.features[x$SCT@var.features %in% pc]
    print(length(sample_var_genes))
    print(length(sample_var_genes))
    all_var_features <- union(all_var_features, sample_var_genes)
}
print(length(all_var_features))

for (x in combined) {
  print(length(rownames(x$SCT)))
    all_var_features <- all_var_features[all_var_features %in% rownames(x$SCT)]
}
print(length(all_var_features))

list_genes <- map(combined, function(x) { x$SCT@var.features[x$SCT@var.features %in% pc] })

# initialize matrix
samples <- names(list_genes)
n <- length(samples)
shared_matrix <- matrix(0, nrow = n, ncol = n, dimnames = list(samples, samples))

# calculate pairwise proportion of shared proteins (Jaccard index)
for (i in seq_along(samples)) {
  for (j in seq_along(samples)) {
    a <- list_genes[[i]]
    b <- list_genes[[j]]
    shared_matrix[i, j] <- length(intersect(a, b)) / length(union(a, b))
  }
}

# plot heatmap again (should be cleaner) 
pheatmap(shared_matrix,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         main = "Proportion of shared proteins")
```

```{r}
# merge list into one object 
combined_merged <- merge(combined[[1]], y = combined[-1])
```

```{r}
# define hvgs (the protein coding genes)
all_var_features
VariableFeatures(combined_merged) <- all_var_features
```

```{r}
# run pca (this is the “integration input”)
combined_merged <- RunPCA(
  combined_merged,
  features = all_var_features,
  npcs = 20,
  verbose = TRUE
)
ElbowPlot(combined_merged)
```

# Integrate
```{r}
# integrating + batch correcting by donor 
combined_merged <- IntegrateLayers(
  object = combined_merged,
  method = HarmonyIntegration,
  orig.reduction = "pca",
  new.reduction = "harmony",
  group.by = "orig.ident"
)
```

# PCA + Neighbors + Clustering 
```{r}
# neighbors + clustering + pca 
combined_merged <- combined_merged %>% 
  RunPCA(verbose = T) %>%
  RunUMAP(dims = 1:20, verbose = T) %>%
  FindNeighbors(dims = 1:20, verbose = T) %>% 
  FindClusters(resolution = c(0.01, 0.05, 0.1, seq(0.2, 1.4, 0.2)))

DimPlot(combined_merged, group.by = c("integrated_snn_res.0.2"), label = T)
```

```{r}
# neighbors + clustering + umap - .1
combined_merged.1 <- RunUMAP(combined_merged, reduction = "harmony", dims = 1:20)
combined_merged.1 <- FindNeighbors(combined_merged, reduction = "harmony", dims = 1:20)
combined_merged.1 <- FindClusters(combined_merged, resolution = c(0.01, 0.05, 0.1, seq(0.2, 1.4, 0.2)))
```

--------------------------------------------------------------------------------
# Load data w/o whole process
```{r}
combined_merged <- readRDS("combined_merged")
```

# Validations of integration
```{r}
# plot by donor before vs after
DimPlot(combined_merged, reduction = "pca", group.by = "orig.ident") + NoLegend() # PCA → cells separate by donor
DimPlot(combined_merged, reduction = "harmony", group.by = "orig.ident") + NoLegend() # Harmony → donors are mixed within clusters

# plot clusters after Harmony
DimPlot(combined_merged, reduction = "harmony", label = TRUE)

# quantify mixing
table(combined_merged$orig.ident, combined_merged$seurat_clusters) # each cluster contains multiple donors (not dominated by one)
# raw cell counts (number of cells from each donor in each cluster)
```

# Visualizations
```{r}
Reductions(combined_merged) # check reductions 
colnames(combined_merged@meta.data) # check metadata

# testing combined_merged
DimPlot(combined_merged, reduction = "umap") # this is the seurat clusters
DimPlot(combined_merged, split.by = "batch") # this is the seurat clusters but split by batch 
DimPlot(combined_merged, reduction = "umap", group.by = "orig.ident") + NoLegend() # should be mixed
DimPlot(combined_merged, reduction = "umap", group.by = "batch") # AUS vs ADCO mixed
DimPlot(combined_merged, reduction = "umap", group.by = "Status") # AD vs control still separable

# testing combined_merged.1
DimPlot(combined_merged.1, reduction = "umap") # seurat clusters 
DimPlot(combined_merged.1, reduction = "umap", group.by = "orig.ident") + NoLegend() # should be mixed
DimPlot(combined_merged.1, reduction = "umap", group.by = "batch") # AUS vs ADCO mixed
DimPlot(combined_merged.1, reduction = "umap", group.by = "Status") # AD vs control still separable

# comparisons 
DimPlot(combined_merged.1, reduction = "umap", group.by = "integrated_snn_res.0.2")
DimPlot(combined_merged.1, reduction = "umap", group.by = "SCT_snn_res.0.2")
```

# Combining sub-status
```{r}
# split by batch, status, & explanatory status (ex: young & old controls)
combined_merged$Status_combined <- ifelse(
  !is.na(combined_merged$Case_sub_status),
  paste0("AUS_", combined_merged$Case_sub_status),
  paste0("P4_", combined_merged$Status2)
)
table(combined_merged$Case_sub_status)
table(combined_merged$Status2)
table(combined_merged$Status_combined)
DimPlot(combined_merged, reduction = "umap", group.by = "Status_combined")
# DimPlot(combined_merged, reduction = "umap_cnmf", group.by = "Status_combined")

# combined all controls together - including each batch 
combined_merged$Status_mixed <- ifelse(
  grepl("CT|CO", combined_merged$Status_combined),
  "Control",
  combined_merged$Status_combined
)
table(combined_merged$Status_mixed)
table(combined_merged$Status_combined[combined_merged$Status_mixed == "Control"])
DimPlot(combined_merged, reduction = "umap", group.by = "Status_mixed")
# DimPlot(combined_merged, reduction = "umap_cnmf", group.by = "Status_mixed")
table(combined_merged$seurat_clusters, combined_merged$Status_mixed) # sub status proportions across all clusters 

# combining AUS_sporadic & P4_AD
combined_merged$Status_final <- ifelse(
  combined_merged$Status_mixed %in% c("AUS_sporadic", "P4_AD"),
  "sAD",
  combined_merged$Status_mixed
)

table(combined_merged$Status_final) 
DimPlot(combined_merged, reduction = "umap", group.by = "Status_final")
# DimPlot(combined_merged, reduction = "umap_cnmf", group.by = "Status_final")

# combined all controls together - split by batch 
combined_merged$Status.3 <- ifelse(
  grepl("CT", combined_merged$Status_combined),
  "AUS_control",
  ifelse(
    grepl("CO", combined_merged$Status_combined),
    "P4_control",
    combined_merged$Status_combined
  )
)
table(combined_merged$Status.3)
DimPlot(combined_merged, reduction = "umap", group.by = "Status.3")
# DimPlot(combined_merged, reduction = "umap_cnmf", group.by = "Status.3")
table(combined_merged$seurat_clusters, combined_merged$Status_mixed) # sub status proportions across all clusters 
```

# Cohort proportions per cluster
```{r}
# quantify mixing - showing the number of cells from that donor in that cluster
table(combined_merged$orig.ident, combined_merged$seurat_clusters) 
sum(table(combined_merged$orig.ident, combined_merged$seurat_clusters)) # making sure that it equals the total number of cells 
ncol(combined_merged) # should match 

# proportion table
tab <- table(combined_merged$batch, combined_merged$seurat_clusters)
prop_tab <- prop.table(tab, margin = 2)  # normalize per cluster
prop_tab # values = fraction of cells from each cohort in that cluster
# observation: cluster 13 is only represented by the Australia data - not by just 1 donor though

# plot 
df <- as.data.frame(prop_tab)
colnames(df) <- c("Batch", "Cluster", "Proportion")
ggplot(df, aes(x = Cluster, y = Proportion, fill = Batch)) +
  geom_bar(stat = "identity") +
  theme_classic()

# check if cluster 13 is AD vs CO vs mixed
table(combined_merged$Status, combined_merged$seurat_clusters)

# proportion table - status_mixed
tab <- table(combined_merged$Status_mixed, combined_merged$seurat_clusters)
prop_tab <- prop.table(tab, margin = 2)  # normalize per cluster
prop_tab # values = fraction of cells from each cohort in that cluster

# plot 
df <- as.data.frame(prop_tab)
colnames(df) <- c("Status_mixed", "Cluster", "Proportion")
ggplot(df, aes(x = Cluster, y = Proportion, fill = Status_mixed)) + geom_bar(stat = "identity") + theme_classic()

# proportion table - status_final
tab <- table(combined_merged$Status_final, combined_merged$seurat_clusters)
prop_tab <- prop.table(tab, margin = 2)  # normalize per cluster
prop_tab # values = fraction of cells from each cohort in that cluster

# plot 
df <- as.data.frame(prop_tab)
colnames(df) <- c("Status_final", "Cluster", "Proportion")
ggplot(df, aes(x = Cluster, y = Proportion, fill = Status_final)) +
  geom_bar(stat = "identity") +
  theme_classic()

# proportion table - status.3
tab <- table(combined_merged$Status.3, combined_merged$seurat_clusters)
prop_tab <- prop.table(tab, margin = 2)  # normalize per cluster
prop_tab # values = fraction of cells from each cohort in that cluster

# plot 
df <- as.data.frame(prop_tab)
colnames(df) <- c("Status.3", "Cluster", "Proportion")
ggplot(df, aes(x = Cluster, y = Proportion, fill = Status.3)) +
  geom_bar(stat = "identity") +
  theme_classic()
```

```{r}
# create directory
# dir.create("downstream/new_filtering/qs/harmony_integrated", recursive = TRUE, showWarnings = FALSE)
qsave(combined_merged, file = "downstream/new_filtering/qs/harmony_integrated/phase4+australia.fromqmd")
```

# Saving files to export for cnmf in python
```{r}
# 1. save HVG gene list
write.table(
  all_var_features,
  file = "microglia_cnmf_input/R_1211_genes.txt",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)

# 2. extract counts matrix
counts <- GetAssayData(combined_merged, assay = "RNA", slot = "counts")
counts <- as(counts, "dgCMatrix")
# sanity check
dim(counts) # should bee 24388 cells 

# 3. save matrix (MTX format)
writeMM(counts, "microglia_cnmf_input/matrix_1211.mtx")

# 4. save genes + barcodes
write.table(
  rownames(counts),
  "microglia_cnmf_input/genes_1211.tsv",
  row.names = FALSE,
  col.names = FALSE,
  quote = FALSE
)

write.table(
  colnames(counts),
  "microglia_cnmf_input/barcodes_1211.tsv",
  row.names = FALSE,
  col.names = FALSE,
  quote = FALSE
)

# 5. save metadata 
write.table(
  combined_merged@meta.data,
  "microglia_cnmf_input/metadata_1211.tsv",
  sep = "\t",
  quote = FALSE,
  col.names = NA   # <-- IMPORTANT (keeps cell IDs aligned)
)

# 6. check files
list.files("microglia_cnmf_input")
```

--------------------------------------------------------------------------------
# Analysis after running cnmf 
```{r}
# loading coordinates 
umap_coords <- read.csv(
  "microglia_batchcorrect/microglia_1211/microglia_umap_1211_coords.csv",
  row.names = 1
)
combined_merged[["umap_cnmf"]] <- CreateDimReducObject(
  embeddings = as.matrix(umap_coords),
  key = "UMAPCNMF_",
  assay = DefaultAssay(combined_merged)
)
```

```{r}
# load cnmf results back into R
data_dir <- "./microglia_batchcorrect/microglia_1211"

# usage → cell-by-program matrix
usage_file <- file.path(
  data_dir,
  "microglia_1211.usages.k_8.dt_0_1.consensus.txt"
)
# spectra_score → gene-by-program scores
spectra_score_file <- file.path(
  data_dir,
  "microglia_1211.gene_spectra_score.k_8.dt_0_1.txt"
)
# spectra_tpm → gene expression per program
spectra_tpm_file <- file.path(
  data_dir,
  "microglia_1211.gene_spectra_tpm.k_8.dt_0_1.txt"
)

usage <- read.table(usage_file, sep="\t", header=TRUE, row.names=1)
spectra_score <- read.table(spectra_score_file, sep="\t", header=TRUE, row.names=1)
spectra_tpm <- read.table(spectra_tpm_file, sep="\t", header=TRUE, row.names=1)

# normalize usage
usage_norm <- t(apply(usage, 1, function(x) x / sum(x))) # This makes each cell sum to 1 so programs become interpretable proportions.
usage_norm <- as.data.frame(usage_norm)

# add usage back into your Seurat object
combined_merged@meta.data <- cbind(
  combined_merged@meta.data,
  usage_norm
)

head(combined_merged@meta.data[, colnames(usage_norm)]) # check if added to metadata
```

```{r}
# plot programs on UMAP
FeaturePlot(
  combined_merged,
  features = colnames(usage_norm),
  reduction = "umap_cnmf",
  combine = FALSE
)

plots <- FeaturePlot(
  combined_merged,
  features = colnames(usage_norm),
  reduction = "umap_cnmf",
  combine = FALSE
)
wrap_plots(plots, ncol = 4)

```
```{r}
# dominant Program per Cell
dominant_program <- apply(usage[,1:8], 1, which.max)
combined_merged$dominant_program <- dominant_program
DimPlot(combined_merged, group.by = "dominant_program", reduction = "umap_cnmf")
```
```{r}
# one UMAP per Dominant Program
dominant_program <- apply(usage[,1:8], 1, which.max) # Determine dominant program
combined_merged$dominant_program <- paste0("Program_", dominant_program)
DimPlot(
  combined_merged,
  cells.highlight = colnames(combined_merged)[combined_merged$dominant_program == "Program_1"]
)
# loop 
p1_cells <- colnames(combined_merged)[combined_merged$dominant_program == "Program_1"]
DimPlot(
  combined_merged,
  cells.highlight = p1_cells
)
for(i in 1:8){
  
  prog_name <- paste0("Program_", i)
  
  cells <- colnames(combined_merged)[combined_merged$dominant_program == prog_name]
  
  print(
    DimPlot(
      combined_merged,
      reduction = "umap_cnmf",
      cells.highlight = cells
    ) + ggtitle(prog_name)
  )
}

```

# Top genes per program 
```{r}
# extract top genes per GEP
get_top_genes <- function(row){
  top_idx <- order(row, decreasing=TRUE)[1:100]
  colnames(spectra_score)[top_idx]
}
top_genes <- apply(spectra_score, 1, get_top_genes)
top_genes <- as.data.frame(top_genes)
top_genes

# saving file for enrichment analysis in r
top_list <- lapply(rownames(spectra_score), function(p){
  get_top_genes_df(p, spectra_score, n = 100) # getting top 100 genes per program 
})
top_genes_df <- do.call(rbind, top_list)

# write to file - saving to export to terminal r for enrichment analysis 
write.csv(
  top_genes_df,
  "combined_merged.1_cnmf_k8_top100_genes.csv",
  row.names = FALSE
)
```

# Plot distribution of genes programs per individual to identify programs more prevalent in AD vs controls
```{r}
# get your usage matrix into metadata
usage <- usage[colnames(combined_merged), ]
combined_merged@meta.data <- cbind(combined_merged@meta.data, usage)
# Add donor + condition info
combined_merged$orig.ident
combined_merged$Status   # "AD" or "Control"
any(duplicated(colnames(combined_merged@meta.data))) # is anything duplicated 
colnames(combined_merged@meta.data)[duplicated(colnames(combined_merged@meta.data))] # what is duplicated 
colnames(usage) <- paste0("Program_", seq_len(ncol(usage))) # Rename program columns to program...
# re-add to metadata
combined_merged@meta.data <- combined_merged@meta.data[, 
                                                   !colnames(combined_merged@meta.data) %in% colnames(usage)
]
combined_merged@meta.data <- cbind(combined_merged@meta.data, usage)
# aggregate to donor level
program_cols <- colnames(usage)
donor_program_usage <- combined_merged@meta.data %>%
  group_by(orig.ident, Status) %>%
  summarise(across(all_of(program_cols), mean))
combined_merged@meta.data[program_cols] <- lapply(
  combined_merged@meta.data[program_cols],
  as.numeric
)

donor_program_usage <- combined_merged@meta.data %>%
  group_by(orig.ident, Status) %>%
  summarise(across(all_of(program_cols), mean), .groups = "drop")
# convert to long format for plotting
plot_df <- donor_program_usage %>%
  pivot_longer(
    cols = all_of(program_cols),
    names_to = "Program",
    values_to = "Usage"
  )
# plot distributions
ggplot(plot_df, aes(x = Status, y = Usage, fill = Status)) +
  geom_boxplot() +
  facet_wrap(~ Program, scales = "free_y") +
  theme_classic()

# statistical testing - to formally identify programs enriched in AD
stats <- plot_df %>%
  group_by(Program) %>%
  wilcox_test(Usage ~ Status) %>%
  adjust_pvalue(method = "BH")

# ploting for status_mixed
# aggregate to donor level
program_cols <- colnames(usage)
donor_program_usage <- combined_merged@meta.data %>%
  group_by(orig.ident, Status_mixed) %>%
  summarise(across(all_of(program_cols), mean))
combined_merged@meta.data[program_cols] <- lapply(
  combined_merged@meta.data[program_cols],
  as.numeric
)

donor_program_usage <- combined_merged@meta.data %>%
  group_by(orig.ident, Status_mixed) %>%
  summarise(across(all_of(program_cols), mean), .groups = "drop")
# convert to long format for plotting
plot_df <- donor_program_usage %>%
  pivot_longer(
    cols = all_of(program_cols),
    names_to = "Program",
    values_to = "Usage"
  )
# plot distributions
ggplot(plot_df, aes(x = Status_mixed, y = Usage, fill = Status_mixed)) +
  geom_boxplot() +
  facet_wrap(~ Program, scales = "free_y") +
  theme_classic()

# statistical Testing - to formally identify programs enriched in AD
stats <- plot_df %>%
  group_by(Program) %>%
  wilcox_test(Usage ~ Status_mixed) %>%
  adjust_pvalue(method = "BH")

# ploting for status_final
# aggregate to donor level
program_cols <- colnames(usage)
donor_program_usage <- combined_merged@meta.data %>%
  group_by(orig.ident, Status_final) %>%
  summarise(across(all_of(program_cols), mean))
combined_merged@meta.data[program_cols] <- lapply(
  combined_merged@meta.data[program_cols],
  as.numeric
)

donor_program_usage <- combined_merged@meta.data %>%
  group_by(orig.ident, Status_final) %>%
  summarise(across(all_of(program_cols), mean), .groups = "drop")
# convert to long format for plotting
plot_df <- donor_program_usage %>%
  pivot_longer(
    cols = all_of(program_cols),
    names_to = "Program",
    values_to = "Usage"
  )
# plot distributions
ggplot(plot_df, aes(x = Status_final, y = Usage, fill = Status_final)) +
  geom_boxplot() +
  facet_wrap(~ Program, scales = "free_y") +
  theme_classic()

# statistical Testing - to formally identify programs enriched in AD
stats <- plot_df %>%
  group_by(Program) %>%
  wilcox_test(Usage ~ Status_final) %>%
  adjust_pvalue(method = "BH")

# ploting for status.3
# aggregate to donor level
program_cols <- colnames(usage)
donor_program_usage <- combined_merged@meta.data %>%
  group_by(orig.ident, Status.3) %>%
  summarise(across(all_of(program_cols), mean))
combined_merged@meta.data[program_cols] <- lapply(
  combined_merged@meta.data[program_cols],
  as.numeric
)

donor_program_usage <- combined_merged@meta.data %>%
  group_by(orig.ident, Status.3) %>%
  summarise(across(all_of(program_cols), mean), .groups = "drop")
# convert to long format for plotting
plot_df <- donor_program_usage %>%
  pivot_longer(
    cols = all_of(program_cols),
    names_to = "Program",
    values_to = "Usage"
  )
# plot distributions
ggplot(plot_df, aes(x = Status.3, y = Usage, fill = Status.3)) +
  geom_boxplot() +
  facet_wrap(~ Program, scales = "free_y") +
  theme_classic()

# statistical testing - to formally identify programs enriched in AD
stats <- plot_df %>%
  group_by(Program) %>%
  wilcox_test(Usage ~ Status.3) %>%
  adjust_pvalue(method = "BH")
```

# Fisher exact test - compared to Anjali's paper - do after gene overap analysis 
```{r}
# asssign each cell a cell state
usage <- combined_merged@meta.data[, program_cols]
# assign dominant program per cell
combined_merged$top_program <- colnames(usage)[max.col(usage, ties.method = "first")]

# fixed mapping (match Program_1 ... Program_8)
program_to_state <- c(
  "Program_1" = "Homeostatic",
  "Program_2" = "IFN-III",
  "Program_3" = "No Overlap",
  "Program_4" = "IFN-I",
  "Program_5" = "DAM",
  "Program_6" = "IFN-II",
  "Program_7" = "MHCII",
  "Program_8" = "No Overlap"   # <-- you MUST define this
)

# assign states (correct way)
cell_states <- program_to_state[combined_merged$top_program]
names(cell_states) <- colnames(combined_merged)
combined_merged$cell_state <- cell_states
table(combined_merged$cell_state, useNA = "ifany") # check

state_props <- combined_merged@meta.data %>%
  group_by(orig.ident, Status, cell_state) %>%
  summarise(n_cells = n(), .groups = "drop") %>%
  group_by(orig.ident) %>%
  mutate(prop = n_cells / sum(n_cells))

stats_states <- state_props %>%
  group_by(cell_state) %>%
  wilcox_test(prop ~ Status) %>%
  adjust_pvalue(method = "BH")

ggplot(state_props, aes(x = Status, y = prop, fill = Status)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1) +
  facet_wrap(~ cell_state, scales = "free_y") +
  theme_classic()

# order the states
combined_merged$cell_state <- factor(
  combined_merged$cell_state,
  levels = c("Homeostatic", "DAM", "IFN-I", "IFN-II", "IFN-III", "MHCII")
)

DimPlot(combined_merged, reduction = "umap_cnmf", label = T) + ggtitle("Cluster Overlay")# seurat clusters - r clusters overlay
DimPlot(combined_merged, group.by = "batch", reduction = "umap_cnmf", label = T) + ggtitle("Batch") # AUS vs ADCO mixed
DimPlot(combined_merged, group.by = "orig.ident", reduction = "umap_cnmf") + NoLegend() + ggtitle("Donor") 
DimPlot(combined_merged, group.by = "Status_final", reduction = "umap_cnmf") + ggtitle("Status Final") 
DimPlot(combined_merged, group.by = "dominant_program", reduction = "umap_cnmf") + ggtitle("Dominant Program")
DimPlot(combined_merged, group.by = "cell_state", reduction = "umap_cnmf", split.by = "Status_final") + ggtitle("Status_final")
VlnPlot(combined_merged, features = "percent.mt") 

DimPlot(combined_merged, group.by = "top_program", reduction = "umap_cnmf", label = T) + ggtitle("Dominant Program") # 8 dominant programs 
DimPlot(combined_merged, group.by = "cell_state", reduction = "umap_cnmf", label = T) + ggtitle("Cell State") # state labels 
DimPlot(combined_merged, group.by = "cell_state", reduction = "umap_cnmf", label = TRUE, repel = TRUE) # state labels 
DimPlot(combined_merged, group.by = "cell_state", reduction = "umap_cnmf", split.by = "batch") + ggtitle("Batch Split") # state label by batch 
DimPlot(combined_merged, group.by = "cell_state", reduction = "umap_cnmf") +
  facet_wrap(~ cell_state) + ggtitle("Cell States")

DimPlot(combined_merged, group.by = "top_program", reduction = "umap_cnmf") +
  facet_wrap(~ top_program) + ggtitle("Dominant Program")

DimPlot(combined_merged, group.by = "cell_state", reduction = "umap_cnmf", split.by = "Status.3") + ggtitle("Status.3 Split") # state label by status.3 
DimPlot(combined_merged, group.by = "top_program", reduction = "umap_cnmf", split.by = "Status.3") + ggtitle("Status.3 Split") # state label by status.3  

# ---- same facet plot wut including the rest of the cells in the plot as grey ----
# extract UMAP coordinates
umap_df <- Embeddings(combined_merged, "umap_cnmf") %>%
  as.data.frame()
umap_df$top_program <- combined_merged$top_program
# Create a plotting dataframe: highlight cells belonging to facet program, everything else stays grey
plot_list <- lapply(unique(umap_df$top_program), function(p) {
  tmp <- umap_df
  tmp$highlight <- ifelse(tmp$top_program == p, p, "Other")
  tmp$facet_program <- p
  tmp
})

plot_df <- bind_rows(plot_list)
program_colors <- c(
  "Program_1" = "#F8766D",
  "Program_2" = "#D89000",
  "Program_3" = "#7CAE00",
  "Program_4" = "#00BA38",
  "Program_5" = "#00BFC4",
  "Program_6" = "#619CFF",
  "Program_7" = "#C77CFF",
  "Program_8" = "#FF61CC",
  "Other" = "grey85"
)

ggplot(plot_df,
       aes(UMAPCNMF_1, UMAPCNMF_2, color = highlight)) +
  geom_point(size = 0.2) +
  facet_wrap(~ facet_program) +
  scale_color_manual(values = program_colors) +
  theme_classic() +
  labs(
    title = "Dominant Program",
    color = NULL
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

colnames(combined_merged@meta.data)

tab <- table(combined_merged$batch, combined_merged$dominant_program)
addmargins(tab)

```

# Number of programs per cell
```{r}
# create histogram showing distribution of programs expressed per cell (1 program, 2 programs, 3 programs... 4,5,6,7,8)
class(usage) # check what the usage matrix is (data-frame)
str(usage)
# convert just the numeric part - to view histogram for picking threshold 
usage_numeric <- as.matrix(usage[,1:8])
hist(usage_numeric)
# look at distribution - a big spike near 0 (noise) / a tail of higher values (real signal) - you want something that separates those 
hist(usage_numeric)
hist(as.vector(usage_numeric)) # same thing

# decide what counts as “expressed” - “Only count a program if it contributes enough to the cell”
threshold <- 0.1

# count programs per cell - Convert usage → binary (expressed or not), then sum per cell:
## usage_matrix = cells x programs
binary_usage <- usage > threshold
programs_per_cell <- rowSums(binary_usage)

# plot histogram
hist(programs_per_cell,
     breaks = 0:max(programs_per_cell),
     main = "Programs per Cell",
     xlab = "Number of Programs",
     ylab = "Number of Cells")

# ggplot2 (better visualization)
df <- data.frame(programs_per_cell = programs_per_cell)

ggplot(df, aes(x = programs_per_cell)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Programs per Cell",
       x = "Number of Programs Expressed",
       y = "Cell Count") +
  theme_minimal()

# discrete bar plot
ggplot(df, aes(x = factor(programs_per_cell))) +
geom_bar(fill = "steelblue") +
  labs(title = "Programs per Cell",
       x = "Number of Programs",
       y = "Number of Cells") +
  theme_minimal()

max(programs_per_cell) # check if any cells express all 7 programs
sum(programs_per_cell == 7) # 0 → none exist
table(programs_per_cell) # look at the full distribution (including zeros) 
table(df$programs_per_cell)
## no cells meet your threshold for all 8 programs ##

# add programs_per_cell back into your Seurat object
combined_merged@meta.data <- cbind(
  combined_merged@meta.data,
  programs_per_cell
)

head(combined_merged@meta.data[, colnames(programs_per_cell)]) # check if added to metadata
```

# Histogram showing distribution of usage ranges for each program
```{r}
# example: df should contain donor_id, cell_id, program, usage
# ---- bin usage into ranges ----
df_hist <- df.2 %>%
  mutate(
    usage_bin = cut(
      usage,
      breaks = seq(0, 1, by = 0.1),   # adjust if needed
      include.lowest = TRUE
    )
  )
# ---- histogram / barplot of usage distribution per program ----
ggplot(df_hist, aes(x = usage_bin, fill = usage_bin)) +
  geom_bar(color = "black") +
  facet_wrap(~ program, scales = "free_y") +
  labs(
    title = "Distribution of Usage Values for Each Program",
    x = "Program Usage Range",
    y = "Number of Cells"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )

# ---- create usage bins ----
df_counts <- df.2 %>%
  mutate(
    usage_bin = cut(
      usage,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE
    )
  ) %>%
  # count cells per program per usage range
  count(program, usage_bin) %>%
  # optional: make table wide format
  tidyr::pivot_wider(
    names_from = usage_bin,
    values_from = n,
    values_fill = 0
  )

df_counts

# ---- continuous histogram instead of discrete bins ----
ggplot(df.2, aes(x = usage)) +
  geom_histogram(
    bins = 30,
    color = "black",
    fill = "steelblue"
  ) +
  facet_wrap(~ program, scales = "free_y") +
  labs(
    title = "Program Usage Distributions",
    x = "Usage",
    y = "Cell Count"
  ) +
  theme_bw()

# ---- histogram of dominant program counts per cell by usage bin ----
# ---- create matching program labels ----
df_dom <- df.2 %>%
  mutate(
    dominant_match = paste0(
      "X",
      gsub("Program_", "", dominant_program)
    )
  ) %>%
  filter(program == dominant_match)

table(df_dom$program) # check
table(df_dom$dominant_program) # check 

df_dom_counts <- df_dom %>%
  mutate(
    usage_bin = cut(
      usage,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE
    )
  ) %>%
  count(dominant_program, usage_bin) %>%
  tidyr::pivot_wider(
    names_from = usage_bin,
    values_from = n,
    values_fill = 0
  )

df_dom_counts

# ---- percentages of df_dom_counts ----
tab <- df_dom_counts
mat <- as.matrix(tab[,-1])
rownames(mat) <- tab$dominant_program

prop_row <- prop.table(mat, margin = 1) * 100
round(prop_row, 1)
# Add row totals
tab_row_total <- cbind(mat, Total = rowSums(mat))
tab_row_total
# Add bottom column totals
tab_full <- rbind(
  tab_row_total,
  Total = c(colSums(mat), sum(mat))
)
tab_full

# Counts + percentages together
prop_row <- prop.table(mat, 1) * 100
combined <- matrix(
  paste0(mat, " (", round(prop_row, 1), "%)"),
  nrow = nrow(mat)
)
rownames(combined) <- rownames(mat)
colnames(combined) <- colnames(mat)
combined
# Add totals to combined table
combined_df <- as.data.frame(combined)
combined_df$Total <- rowSums(mat)
bottom_row <- c(colSums(mat), sum(mat))
combined_df <- rbind(combined_df, Total = bottom_row)
combined_df

prop.table(mat, 1)

# ---- histogram of dominant program ----
programs <- unique(df_dom$dominant_program)

for (p in programs) {

  df_p <- df_dom %>%
    filter(dominant_program == p)

  p1 <- ggplot(df_p, aes(x = usage)) +
    geom_histogram(
      bins = 30,
      fill = "steelblue",
      color = "black"
    ) +
    labs(
      title = paste("Usage Distribution for Dominant Program:", p),
      x = "Usage",
      y = "Number of Cells"
    ) +
    theme_bw()

  print(p1)
}

# another histogram - aligned with the counts by 10s 
head(df_hist$program)
head(df_hist$dominant_program)

df_hist <- df_hist %>%
  mutate(
    program = sub("^X", "Program_", program)
  )

df_dom_hist <- df_hist %>%
  filter(program == dominant_program)

nrow(df_dom_hist) # should be all cells 
table(df_dom_hist$program)
head(df_hist$program)
head(df_hist$dominant_program)

ggplot(df_dom_hist, aes(x = usage_bin, fill = usage_bin)) +
  geom_bar(color = "black") +
  facet_wrap(~ program, scales = "free_y") +
  labs(
    title = "Distribution of Usage Values for Dominant Cells per Program",
    x = "Program Usage Range",
    y = "Number of Cells"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )
# question: for the cells that have low usage but still are dominate for a program, i want to see what to other usage ratios are for the other program. just bc the usage is low but yet still dominant. 
```

# Line graph showing programs and their usage ranges
```{r}
# ---- bin usage values ----
df_line <- df.2 %>%
  mutate(
    usage_bin = cut(
      usage,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE
    )
  ) %>%
  # count cells in each usage range for each program
  count(program, usage_bin)

# ---- line graph ----
ggplot(df_line,
       aes(x = usage_bin,
           y = n,
           group = program,
           color = program)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Program Usage Range Distributions Across All Cells",
    x = "Usage Range",
    y = "Number of Cells"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# ---- facet line graphs ----
ggplot(df_line,
       aes(x = usage_bin,
           y = n,
           group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ program, scales = "free_y") +
  labs(
    title = "Usage Range Distribution per Program Across All Cells",
    x = "Usage Range",
    y = "Number of Cells"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )

# ---- non facet line graph ----
df_bin <- df.2 %>%
  mutate(
    usage_bin = cut(
      usage,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE
    )
  ) %>%
  count(program, usage_bin)
programs <- unique(df_bin$program)
for (p in programs) {
  df_p <- df_bin %>% filter(program == p)
  print(
    ggplot(df_p, aes(x = usage_bin, y = n, group = 1)) +
      geom_line(linewidth = 1) +
      geom_point(size = 2) +
      labs(
        title = paste("Usage Distribution Across All Cells -", p),
        x = "Usage Range",
        y = "Number of Cells"
      ) +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}

# ---- line graph of dominant program ----
programs <- unique(df_dom$dominant_program)
for (p in programs) {

  df_p <- df_dom %>%
    filter(dominant_program == p)

  p2 <- ggplot(df_p, aes(x = usage)) +
    geom_density(fill = "lightblue", alpha = 0.5) +
    labs(
      title = paste("Density Distribution:", p),
      x = "Usage",
      y = "Density"
    ) +
    theme_bw()

  print(p2)
}
```

# Box plot comparing programs per cell by status (ADAD, sporadic, controls), correcting for age and other covariates. 
```{r}
ggplot(df.2, aes(x = Status.3, y = programs_per_cell)) +
  geom_boxplot(outlier.shape = NA) +   # cleaner look (hides extreme outliers)
  geom_jitter(width = 0.2, alpha = 0.2, size = 0.8) +  # shows individual cells
  theme_classic() +
  labs(
    x = "Status",
    y = "# Programs per Cell"
  )

ggplot(df.2, aes(x = Status.3, y = programs_per_cell, fill = Status.3)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.15, size = 0.7) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(
    x = "Disease Status",
    y = "# Programs per Cell"
  )

# box plot comparing programs per cell by status - before covariate correction 
ggplot(df.2, aes(x = Status.3, y = programs_per_cell, fill = Status.3)) +
  geom_boxplot() +
  theme_classic()

# box plot comparing programs per cell by status - CORRECTING FOR COVARIATES
adj_model <- lm(programs_per_cell ~ status + age + sex + batch, data = df.2)
df.2$adj_programs <- resid(adj_model) + mean(df.2$programs_per_cell)
# plot adjusted values
ggplot(df.2, aes(x = Status.3, y = adj_programs, fill = Status.3)) +
  geom_boxplot() +
  theme_classic()

# discrete bar plot - number of programs expressed - combined data - adjusted for covariates - no change 
ggplot(df, aes(x = programs_per_cell)) +
  geom_histogram(binwidth = 1, fill = "grey", color = "black") +
  labs(title = "Distribution of Programs per Cell",
       x = "Number of Programs Expressed",
       y = "Cell Count") +
  theme_minimal()

table(df.2$adj_programs) # no change 
```

# Visualizations of programs per cell - which combinations of programs co-occur
```{r}
# step 1: define “program expressed”
threshold <- 0.1
binary_usage <- usage > threshold

# step 2: count programs per cell
program_counts <- rowSums(binary_usage)

# step 3: subset cells with exactly 3 programs
cells_5prog <- binary_usage[program_counts == 3, ]

# step 4: extract actual program combinations
combo_list <- apply(cells_5prog, 1, function(x) {
  paste(names(x)[x == 1], collapse = "_")
})
table(combo_list)

# step 5: heatmap with clustering - Simple heatmap
mat <- binary_usage[program_counts == 3, ]
# Convert TRUE/FALSE → 1/0
mat_numeric <- mat * 1

pheatmap(
  mat_numeric,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE
)

# use continuous values
pheatmap(
  usage[program_counts == 3, ],
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE,
  scale = "row"
)

# step 7: cluster first, then summarize - Hierarchical clustering
hc <- hclust(dist(usage[program_counts == 3, ]))
clusters <- cutree(hc, k = 4)  # k = the number of clusters you choose to divide your cells into
table(clusters) # cluster sizes 
# average program usage per cluster
aggregate(usage[program_counts == 3, ], by = list(clusters), mean)

# heatmap of cluster averages
## one step further and getting a visual of the aggregated data (showing the 4 clear program usage patterns)
cluster_avg <- aggregate(
  usage[program_counts == 3, ],
  by = list(clusters),
  mean
)

rownames(cluster_avg) <- paste0("Cluster_", cluster_avg$Group.1)
cluster_avg$Group.1 <- NULL
pheatmap(
  as.matrix(cluster_avg),
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  scale = "row"
)

# barplot per cluster
barplot(
  t(as.matrix(cluster_avg)),
  beside = TRUE,
  legend = TRUE,
  col = rainbow(7)
)

# dot plot
# add cluster as a column
cluster_avg$Cluster <- rownames(cluster_avg)
# melt properly
df <- melt(cluster_avg, id.vars = "Cluster")
# rename columns
colnames(df) <- c("Cluster", "Program", "Value")
ggplot(df, aes(x = Program, y = Cluster, size = Value, color = Value)) +
  geom_point() +
  scale_color_viridis_c() +
  theme_minimal()
```

# Associate programs with AD vs. control 
```{r}
# step 1: merge usage with metadata
usage_df <- as.data.frame(usage)
usage_df$cell_id <- rownames(usage_df)
meta <- combined_merged@meta.data
meta$cell_id <- rownames(meta)
df <- merge(usage_df, meta, by = "cell_id")

# step 2: compare program usage by condition
program_means <- df %>%
  group_by(Status.3) %>%
  summarise(across(starts_with("GEP"), mean))
# idk this step yet #
df_long <- tidyr::pivot_longer(df,
                               cols = where(is.numeric),
                               names_to = "program",
                               values_to = "usage")

# bar plot of counts by condition
df$programs_per_cell <- programs_per_cell[match(df$cell_id, names(programs_per_cell))]
df$Status.3  # "AD" or "Control" # check
# create a histogram / bar plot of program counts per cell
ggplot(df, aes(x = factor(programs_per_cell), fill = Status.3)) +
  geom_bar(position = "dodge") +  # side-by-side bars
  labs(title = "Distribution of Programs per Cell by Condition",
       x = "Number of Programs Expressed per Cell",
       y = "Cell Count") +
  theme_minimal()

# proportion plot
ggplot(df, aes(x = factor(programs_per_cell, levels = 1:7), fill = Status.3)) +
  geom_bar(position = "fill") +  # normalize to percent
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Cells Expressing Programs by Condition",
       x = "Number of Programs Expressed per Cell",
       y = "Percent of Cells") +
  theme_minimal()

# statistical test
tab <- table(df$programs_per_cell, df$Status)
chisq.test(tab)

# ~~~ status_final ~~~ #
# step 1: merge usage with metadata
usage_df <- as.data.frame(usage)
usage_df$cell_id <- rownames(usage_df)
meta <- combined_merged@meta.data
meta$cell_id <- rownames(meta)
df <- merge(usage_df, meta, by = "cell_id")

# step 2: compare program usage by condition
program_means <- df %>%
  group_by(Status_final) %>%
  summarise(across(starts_with("GEP"), mean))
# idk this step yet #
df_long <- tidyr::pivot_longer(df,
                               cols = where(is.numeric),
                               names_to = "program",
                               values_to = "usage")

# bar plot of counts by condition
df$programs_per_cell <- programs_per_cell[match(df$cell_id, names(programs_per_cell))]
df$Status_final  # "AD" or "Control" # check
# create a histogram / bar plot of program counts per cell
ggplot(df, aes(x = factor(programs_per_cell), fill = Status_final)) +
  geom_bar(position = "dodge") +  # side-by-side bars
  labs(title = "Distribution of Programs per Cell by Condition",
       x = "Number of Programs Expressed per Cell",
       y = "Cell Count") +
  theme_minimal()

Status_final <- combined_merged$Status_final

# proportion plot
# create labels with counts
count_labels <- c(
  "1\n(n=225)",
  "2\n(n=2617)",
  "3\n(n=8539)",
  "4\n(n=9074)",
  "5\n(n=3556)",
  "6\n(n=369)",
  "7\n(n=8)"
)

ggplot(df, aes(x = factor(programs_per_cell), fill = Status_final)) +
  geom_bar(position = "dodge") +
  scale_x_discrete(labels = count_labels) +
  labs(
    title = "Distribution of Programs per Cell by Condition",
    x = "Number of Programs Expressed per Cell",
    y = "Cell Count"
  ) +
  theme_minimal()

colSums(table(df$Status_final, df$programs_per_cell)) # total cells per programs
table(df$Status_final, df$programs_per_cell) # exact cell counts per group × program number

# statistical test
tab <- table(df$programs_per_cell, df$Status_final)
chisq.test(tab)
```

# Investigating cluster 13 
```{r}
table(combined_merged$seurat_clusters)
table(combined_merged$seurat_clusters == 13)

df <- combined_merged@meta.data %>%
     filter(!is.na(Status_mixed)) %>%
     group_by(orig.ident, Status_mixed) %>%
     summarise(prop_13 = mean(cluster13), .groups = "drop")

ggplot(df, aes(x = Status_mixed, y = prop_13, fill = Status_mixed)) +
     geom_boxplot() +
     geom_jitter(width = 0.1) +
     theme_classic()

table(combined_merged$Status_combined[combined_merged$seurat_clusters == 13])
table(combined_merged$Status_mixed[combined_merged$seurat_clusters == 13]) # proportions of sub statuses for cluster 13
prop.table(
  table(combined_merged$Status_mixed[combined_merged$seurat_clusters == 13])
) * 100 # percentage 
wilcox_test(prop_13 ~ Status_mixed, data = df)
as.data.frame(
  table(combined_merged$Status_mixed[combined_merged$seurat_clusters == 13])
) # dataframe of proportions of sub statuses for cluster 13
DimPlot(combined_merged, reduction = "umap", cells.highlight = WhichCells(combined_merged, idents = 13)) # showing where cluster 13 is

# project cluster 13 labels onto the Harmony UMAP coordinates
DimPlot(
  combined_merged,
  reduction = "umap_cnmf",   # your Harmony UMAP
  cells.highlight = WhichCells(combined_merged, idents = 13)
)
```

# All cluster labels projected onto the Harmony umap
```{r}
# currently overlaying old clusters onto a new map
combined_merged$seurat_clusters # check suerat cluster labels 

DimPlot(
  combined_merged,
  reduction = "umap_cnmf",
  group.by = "seurat_clusters",
  label = TRUE
)

DimPlot(
  combined_merged,
  reduction = "umap_cnmf",
  label = TRUE
)
```

# Investigating programs 3 & 8
```{r}
table(combined_merged$Status_final)
table(combined_merged$top_program)
table(
  combined_merged$top_program,
  combined_merged$Status_final
)[c("Program_3", "Program_8"), ]

# check cell counts for programs 3 and 8 across disease groups
combined_merged@meta.data %>%
  filter(top_program %in% c("Program_3", "Program_8")) %>%
  group_by(top_program, Status_final) %>%
  summarise(cell_count = n(), .groups = "drop")

table(
  combined_merged$top_program,
  combined_merged$batch
)[c("Program_3", "Program_8"), ] # programs 3 and 8 are present in both datasets
```

# Status.3 Analysis - by batch and status 
```{r}
table(combined_merged$Status.3)
table(combined_merged$Status.3, combined_merged$dominant_program)
table(combined_merged$Status.3, combined_merged$cell_state)
# create counts table
tab <- table(combined_merged$Status.3, combined_merged$dominant_program)
tab
# convert to row percentages
prop_tab <- prop.table(tab, margin = 2) * 100
round(prop_tab, 1)
# Add row totals (right side)
row_totals <- rowSums(tab)
# Add column totals (bottom row)
col_totals <- colSums(tab)
# Combine everything into one display table
tab_with_row_totals <- cbind(tab, Total = rowSums(tab))
tab_with_col_totals <- rbind(tab_with_row_totals, Total = c(colSums(tab), sum(tab)))
tab_with_col_totals

# percentages + counts together (ex: 707 (12.3%))
prop_tab <- prop.table(tab, margin = 2) * 100
combined <- paste0(tab, " (", round(prop_tab, 1), "%)")
combined

prop.table(tab, 2) # raw decimals, same as percentages 
```

